Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-15 _paper.rmd
-Add ref to fig2/3 in sect 4 text. -obs-> inst, var -> feat - chocolate_inverse .png, and new .mp4 manual capture for all classification.

<!-- Approximations of the spaces, PC1:2 -->
A visualization of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided. A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. <!-- Interactions --> For classification tasks, misclassified instances are circled in red if applicable. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model.

While the comparison of these spaces is interesting, the main purpose of the global view is to enable the selection of instances to explore the local explanations. The projection attribution of the Primary Instance (PI) is examined and typically viewed with an optional Comparison Instance (CI).

@@

<!-- A function of the obs selected -->
The global view facilitated the selection instances. The feature-level attribution of the primary instance is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear features that best explain the difference between the model intercept and an instance's prediction, not the local shape of the model surface.

<!-- Tour start frame -->
The initial frame is the normalized SHAP values of the PI. The distribution normalized values from all instances are shown depicted in the basis as vertical parallel coordinate plots, where each line connects an instance's feature attribution. The PI is highlighted as a dashed line and CI as  a dotted line. The current projection basis is depicted as the width of a bar, the feature contributions to the horizontal axis.

<!-- Tour animation -->
The radial tour creates a basis path by varying the contribution of a selected feature, fully into and out of a projection frame. Doing so tests an individual feature's sensitivity to the structure identified by the local explanation. The default feature selected has the largest discrepancy between the attribution of primary and comparison instances. The following sections elaborate on the takeaways from applying this approach in classification and regression tasks. <!-- Segue to classification -->Now that we have introduced the global view and corresponding cheem radial tour, let us discuss the differences between the classification and regression cases.

@@

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 for data space, attribution space, and prediction by observed y (visual of the confusion matrix for classification). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (c) select features to include, which variable is changed in the tour, and the presence of the parrelle coordiate lines connecting attribution distribution. The visual depicting 1D basis (d) shows the distribution of the feature attribution, bars show the current projection with the black bar is being varied. Panel (e) is the resulting projection of the data shown as density in the classification case."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

The Palmer penguins data [@gorman_ecological_2014; @horst_palmerpenguins_2020] was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to substitute for the overly-used iris data and is quite similar in form. After removing incomplete observations, there are 333 instances and we will use the four physical measurements, `bill_length_mm` (`b_l`), `bill_depth_mm` (`b_d`), `flipper_length_mm` (`f_l`), `body_mass_g` (`wgt`), for this illustration. A random forest model was fit with species as the response feature.

@@

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identified by SHAP in the (wrong) prediction for this instance is mostly `b_l` and `b_d` with small contributions of `f_l` and `wgt`. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap, that is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this, we use the radial tour to vary the contribution of flipper length (`f_l`). (In our exploration, this was the third variable explored. It is typically useful to explore the variables with larger contributions, here `b_l` and `b_d`, but when doing this, nothing was revealed about how the PI differed from other Gentoos.) On varying `f_l` as it contributes more to the projection (b, right), we see that more, and more, this penguin looks like a Chinstrap. This suggests that `f_l` should be considered an important feature for explaining the (wrong) prediction.

Figure \@ref(fig:casepenguinsblfl) confirms that flipper length (`f_l`) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two variables, mostly because it has an unusually low value of flipper length relative to other Gentoos.

@@

The chocolates dataset consists of 88 observations of ten nutritional measurements determined from their labels and labeled as either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University  class STAT503 from nutritional information from the manufacturer's website, and normalized to 100g equivalents. The data is available in the `cheem` package. A random forest model is used for the classification of chocolate type.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate that is nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which of the nutritional variables contribute most to misclassification.

(ref:casechocolates) Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove a 4 variables with the lowest contribution for the selected instance and vary the contribution of `fiber`. The misclassification of two instances seems improbable when sugar is near the max contribution. Animated tour can be found at [shorturl.at/dBE28](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4).
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

This type of exploration is shown in Figure \@ref(fig:casechocolates), where a chocolate labeled dark but predicted to be milk is chosen as the PI (observation 22). It is compared with a CI that is a correctly classified dark chocolate (observation 7). The PCA plot, and the SHAP PCA plots (a) show a big difference between the two chocolate types but with confusion for a handful of observations. The misclassifications are clearer in the observed vs predicted plot, and can be seen to be mistaken in both ways: milk to dark and dark to milk.

The attribution projection for chocolate 22 suggests that Fiber, Sugars and Calories are most responsible for its incorrect prediction. The way to read this plot is to see that Fiber has a large negative value, while Sugars and Calories have reasonably large positive values. In the density plot, observations on the very left of the display would have high values of Fiber (matching the negative projection coefficient) and low values of Sugars and Calories. The opposite would be the interpretation of a point with high values in this plot. The dark chocolates (orange) are mostly on the left, and this is a reason why they are considered to be healthier: high fiber and low sugar. The density for milk chocolates is further to the right, indicating that they generally have low fiber and high sugar.

The observation of interest (dashed line) can be viewed against the comparison observation (dotted line). Now one needs to pay different attention to the parallel plot of the SHAP values, which are local to a particular observation, and the density plot, which is the same projection of all observations as specified by the SHAP values of the observation of interest.

We can quickly compare the variable contributions to the two different predictions from the parallel coordinate plot. The observation of interest differs with the comparison primarily on the Fiber variable, which suggests that this is the reason for the incorrect prediction.

From the density plot, which is the attribution projection corresponding to the observation of interest, both observations are more like dark chocolates. If we vary the contribution of Sugars, and completely remove Sugars from the projection, this is where the difference becomes apparent. When primarily Fiber is examined, observation 22 looks more like a milk chocolate.

<!-- Figure \@ref(fig:casechocolates) similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of `fiber`. We see notice two instances that have clearly well within the milk chocolate despite being observed as dark chocolate on their product label. While conjecture, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. In the penguin case we explored how the model misclassified an observation from a type one error perspective. In contrast, in the chocolates cases shows a type two error view; looking for extreme support of the models classification to challenge the original observation.-->

>>>>

<!-- Approximations of the spaces, PC1:2 -->
A visualization of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided (Fig. \@ref(fig:classificationcase), b) A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. <!-- Interactions --> For classification tasks, misclassified instances are circled in red if applicable. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Fig. \@ref(fig:classificationcase) a and c).

While the comparison of these spaces is interesting, the main purpose of the global view is to enable the selection of instances to explore the local explanations. The projection attribution of the Primary Instance (PI) is examined and typically viewed with an optional Comparison Instance (CI). These instances are are highlighted as asterisk and $\times$ in 2D spaces and long dashed and short dotted lines on 1D spaces.

@@

<!-- Segue of PI to attribution projection -->
The global view facilitated the selection of instances. The attribution projection of the PI is the initial 1D basis in a radial tour. This is an approximation of the feature importance for the prediction of the instance that best explain the difference between the data mean and an instance's prediction, not the local shape of the model surface.

<!-- Tour start frame -->
Normalized attributions of the PI and CI are shown as dashed and dotted lines against the distribution of the attribution from all instances. These are depicted as a vertical parallel coordinate plot, where each line connects one instance's feature attribution (Fig. \@ref(fig:classificationcase), d). This is an important global context, the summarization of the attribution of all instances.

<!-- Tour animation -->
The current projection basis is depicted as the width of a bar, the feature contributions to the horizontal positions. The radial tour is going to vary one selected feature to a full contribution, zero contribution, and back to the attribution projection. Doing so tests a feature's sensitivity to the structure identified by the local explanation. The default feature selected has the largest discrepancy between the attribution of the PI and CI. The data (scaled by standard deviation away from the mean) is projected through the current basis --- the horizonal position of Fig. \@ref(fig:classificationcase) e and Fig. \@ref(fig:regressioncase) d.

The following sections elaborate on the difference in displays from applying this approach in classification and regression tasks. Now that we have introduced the global view and corresponding radial tour, let us discuss the differences between the classification and regression cases.

@@

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 for data space, attribution space, and prediction by observed y (visual of the confusion matrix for classification). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (c) select features to include, which feature is changed in the tour, and the presence of the parallel coordiate lines connecting attribution distribution. The visual depicting 1D basis (d) shows the distribution of the feature attribution, bars show the current projection with the black bar is being varied. Panel (e) is the resulting projection of the data shown as density in the classification case."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

The Palmer penguins data [@gorman_ecological_2014; @horst_palmerpenguins_2020] was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to substitute for the overly-used iris data and is quite similar in form. After removing incomplete instances, there are 333 instances and we will use the four physical measurements, `bill_length_mm` (`b_l`), `bill_depth_mm` (`b_d`), `flipper_length_mm` (`f_l`), `body_mass_g` (`wgt`), for this illustration. A random forest model was fit with species as the response feature.

@@

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identified by SHAP in the (wrong) prediction for this instance is mostly `b_l` and `b_d` with small contributions of `f_l` and `wgt`. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap, that is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this, we use the radial tour to vary the contribution of flipper length (`f_l`). (In our exploration, this was the third feature explored. It is typically useful to explore the features with larger contributions, here `b_l` and `b_d`, but when doing this, nothing was revealed about how the PI differed from other Gentoos). On varying `f_l` as it contributes more to the projection (b, right), we see that more, and more, this penguin looks like a Chinstrap. This suggests that `f_l` should be considered an important feature for explaining the (wrong) prediction.

Figure \@ref(fig:casepenguinsblfl) confirms that flipper length (`f_l`) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two features, mostly because it has an unusually low value of flipper length relative to other Gentoos.

@@

The chocolates dataset consists of 88 instances of ten nutritional measurements determined from their labels and labeled as either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University  class STAT503 from nutritional information from the manufacturer's website, and normalized to 100g equivalents. The data is available in the `cheem` package. A random forest model is used for the classification of chocolate type.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate that is nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which of the nutritional features contribute most to misclassification.

(ref:casechocolates) Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove four features with the lowest contribution for the selected instance and vary the contribution of `fiber`. The misclassification of two instances seems improbable when sugar is near the max contribution. Animated tour can be found at [shorturl.at/dBE28](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4).
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

This type of exploration is shown in Figure \@ref(fig:casechocolates), where a chocolate labeled dark but predicted to be milk is chosen as the PI (instance 22). It is compared with a CI that is a correctly classified dark chocolate (instance 7). The PCA plot, and the SHAP PCA plots (a) show a big difference between the two chocolate types but with confusion for a handful of instances. The misclassifications are clearer in the observed vs predicted plot, and can be seen to be mistaken in both ways: milk to dark and dark to milk.

The attribution projection for chocolate 22 suggests that Fiber, Sugars and Calories are most responsible for its incorrect prediction. The way to read this plot is to see that Fiber has a large negative value, while Sugars and Calories have reasonably large positive values. In the density plot, instances on the very left of the display would have high values of Fiber (matching the negative projection coefficient) and low values of Sugars and Calories. The opposite would be the interpretation of a point with high values in this plot. The dark chocolates (orange) are mostly on the left, and this is a reason why they are considered to be healthier: high fiber and low sugar. The density for milk chocolates is further to the right, indicating that they generally have low fiber and high sugar.

The instance of interest (dashed line) can be viewed against the comparison instance (dotted line). Now one needs to pay different attention to the parallel plot of the SHAP values, which are local to a particular instance, and the density plot, which is the same projection of all instances as specified by the SHAP values of the instance of interest.

We can quickly compare the feature contributions to the two different predictions from the parallel coordinate plot. The instance of interest differs with the comparison primarily on the Fiber feature, which suggests that this is the reason for the incorrect prediction.

From the density plot, which is the attribution projection corresponding to the instance of interest, both instances are more like dark chocolates. If we vary the contribution of Sugars, and completely remove Sugars from the projection, this is where the difference becomes apparent. When primarily Fiber is examined, instance 22 looks more like a milk chocolate.

<!-- Figure \@ref(fig:casechocolates) similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of `fiber`. We see notice two instances that have clearly well within the milk chocolate despite being observed as dark chocolate on their product label. While conjecture, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. In the penguin case we explored how the model misclassified an instance from a type one error perspective. In contrast, in the chocolates cases shows a type two error view; looking for extreme support of the models classification to challenge the original instance.-->

### 2022-01-15 _paper.tex
-Add ref to fig2/3 in sect 4 text. -obs-> inst, var -> feat - chocolate_inverse .png, and new .mp4 manual capture for all classification.

SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance, by examining the effect of other features on the predictions. The explanations of SHAP almost all refer to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by assessing the performance of this player in the presence or absence of other players. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. The SHAP values are the mean contributions over some subset of permutations of features. The approach is related to partial dependence plots (Molnar 2020), used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each feature in the context of all other candidate features.

For our application, we use \emph{tree SHAP}, a variant of SHAP that enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over permutations of the feature order, tree SHAP approximates instance-level feature importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized SHAP values as a projection basis (call this the \emph{attribution projection}) which will have coefficients varied in order to further scrutinize the feature contributions.

@@

\caption{Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) is the break-down plot showing one additive feature-attribution.  Panel (b) shows the distribution of attribution for each feature across 25 permutations of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

@@

A visualization of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided. A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. For classification tasks, misclassified instances are circled in red if applicable. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model.

While the comparison of these spaces is interesting, the main purpose of the global view is to enable the selection of instances to explore the local explanations. The projection attribution of the Primary Instance (PI) is examined and typically viewed with an optional Comparison Instance (CI).

@@

The global view facilitated the selection instances. The feature-level attribution of the primary instance is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear features that best explain the difference between the model intercept and an instance's prediction, not the local shape of the model surface.

The initial frame is the normalized SHAP values of the PI. The distribution normalized values from all instances are shown depicted in the basis as vertical parallel coordinate plots, where each line connects an instance's feature attribution. The PI is highlighted as a dashed line and CI as a dotted line. The current projection basis is depicted as the width of a bar, the feature contributions to the horizontal axis.

The radial tour creates a basis path by varying the contribution of a selected feature, fully into and out of a projection frame. Doing so tests an individual feature's sensitivity to the structure identified by the local explanation. The default feature selected has the largest discrepancy between the attribution of primary and comparison instances. The following sections elaborate on the takeaways from applying this approach in classification and regression tasks. Now that we have introduced the global view and corresponding cheem radial tour, let us discuss the differences between the classification and regression cases.

@@

\caption{Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 for data space, attribution space, and prediction by observed y (visual of the confusion matrix for classification). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (c) select features to include, which variable is changed in the tour, and the presence of the parrelle coordiate lines connecting attribution distribution. The visual depicting 1D basis (d) shows the distribution of the feature attribution, bars show the current projection with the black bar is being varied. Panel (e) is the resulting projection of the data shown as density in the classification case.}\label{fig:classificationcase}
\end{figure}

@@

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to substitute for the overly-used iris data and is quite similar in form. After removing incomplete observations, there are 333 instances and we will use the four physical measurements, \texttt{bill\_length\_mm} (\texttt{b\_l}), \texttt{bill\_depth\_mm} (\texttt{b\_d}), \texttt{flipper\_length\_mm} (\texttt{f\_l}), \texttt{body\_mass\_g} (\texttt{wgt}), for this illustration. A random forest model was fit with species as the response feature.

@@

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identified by SHAP in the (wrong) prediction for this instance is mostly \texttt{b\_l} and \texttt{b\_d} with small contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap, that is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this, we use the radial tour to vary the contribution of flipper length (\texttt{f\_l}). (In our exploration, this was the third variable explored. It is typically useful to explore the variables with larger contributions, here \texttt{b\_l} and \texttt{b\_d}, but when doing this, nothing was revealed about how the PI differed from other Gentoos.) On varying \texttt{f\_l} as it contributes more to the projection (b, right), we see that more, and more, this penguin looks like a Chinstrap. This suggests that \texttt{f\_l} should be considered an important feature for explaining the (wrong) prediction.

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{f\_l}) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two variables, mostly because it has an unusually low value of flipper length relative to other Gentoos.

@@

The chocolates dataset consists of 88 observations of ten nutritional measurements determined from their labels and labeled as either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University class STAT503 from nutritional information from the manufacturer's website, and normalized to 100g equivalents. The data is available in the \texttt{cheem} package. A random forest model is used for the classification of chocolate type.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate that is nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which of the nutritional variables contribute most to misclassification.

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove a 4 variables with the lowest contribution for the selected instance and vary the contribution of \texttt{fiber}. The misclassification of two instances seems improbable when sugar is near the max contribution. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolates}
\end{figure}

This type of exploration is shown in Figure \ref{fig:casechocolates}, where a chocolate labeled dark but predicted to be milk is chosen as the PI (observation 22). It is compared with a CI that is a correctly classified dark chocolate (observation 7). The PCA plot, and the SHAP PCA plots (a) show a big difference between the two chocolate types but with confusion for a handful of observations. The misclassifications are clearer in the observed vs predicted plot, and can be seen to be mistaken in both ways: milk to dark and dark to milk.

The attribution projection for chocolate 22 suggests that Fiber, Sugars and Calories are most responsible for its incorrect prediction. The way to read this plot is to see that Fiber has a large negative value, while Sugars and Calories have reasonably large positive values. In the density plot, observations on the very left of the display would have high values of Fiber (matching the negative projection coefficient) and low values of Sugars and Calories. The opposite would be the interpretation of a point with high values in this plot. The dark chocolates (orange) are mostly on the left, and this is a reason why they are considered to be healthier: high fiber and low sugar. The density for milk chocolates is further to the right, indicating that they generally have low fiber and high sugar.

The observation of interest (dashed line) can be viewed against the comparison observation (dotted line). Now one needs to pay different attention to the parallel plot of the SHAP values, which are local to a particular observation, and the density plot, which is the same projection of all observations as specified by the SHAP values of the observation of interest.

We can quickly compare the variable contributions to the two different predictions from the parallel coordinate plot. The observation of interest differs with the comparison primarily on the Fiber variable, which suggests that this is the reason for the incorrect prediction.

From the density plot, which is the attribution projection corresponding to the observation of interest, both observations are more like dark chocolates. If we vary the contribution of Sugars, and completely remove Sugars from the projection, this is where the difference becomes apparent. When primarily Fiber is examined, observation 22 looks more like a milk chocolate.

>>>>

SHaply Additive exPlanations (SHAP) approximates the feature importance in the vicinity of one instance conceptually by examining the effect of other features on the contribution of the feature of interest on predicting the response. These explanations almost all point to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by permuting the players that contributing to the score. Strumbelj and Kononenko (2010) introduced SHAP for local explanations in ML models. While the models themselves are invariant to the order of the predictors, the additive explanations are not. The SHAP values are the mean contributions over some subset of these permutations. This is also related to partial dependence plots (Molnar 2020), used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each feature in the context of all other candidate features.

For our application, we use \emph{tree SHAP}, a variant of SHAP that enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over permutations of the feature order, tree SHAP approximates instance-level feature importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models; we illustrate random forests. The following section will use normalized explanations as to the starting projection basis (call this the \emph{attribution projection}) to further scrutinize the explanation.

@@

\caption{Illustration of the use of SHAP values for a random forest model for salaries of FIFA 2020 players based on nine predictors corresponding to different skills. A star offensive and defensive player, Messi and a defensive player are compared, L. Messi and V van Dijk respectively. Panel (a) is the break-down plot showing one additive feature-attribution. These attributions are sensitive to the order of the features. Panel (b) shows the distribution of attribution for each feature across 25 such permutation of the order of predictors, with the mean displayed as a dot, for each players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

@@

A visualization of these spaces is provided by the first two principal components of their respective spaces. In addition, a plot observed by predicted response is also provided (Fig. \ref{fig:classificationcase}, b) A single 2D projection will not encompass all of the structure of higher-dimensional space, but it is generally a useful visual summary. For classification tasks, misclassified instances are circled in red if applicable. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Fig. \ref{fig:classificationcase} a and c).

While the comparison of these spaces is interesting, the main purpose of the global view is to enable the selection of instances to explore the local explanations. The projection attribution of the Primary Instance (PI) is examined and typically viewed with an optional Comparison Instance (CI). These instances are are highlighted as asterisk and \(\times\) in 2D spaces and long dashed and short dotted lines on 1D spaces.

@@

The global view facilitated the selection of instances. The attribution projection of the PI is the initial 1D basis in a radial tour. This is an approximation of the feature importance for the prediction of the instance that best explain the difference between the data mean and an instance's prediction, not the local shape of the model surface.

Normalized attributions of the PI and CI are shown as dashed and dotted lines against the distribution of the attribution from all instances. These are depicted as a vertical parallel coordinate plot, where each line connects one instance's feature attribution (Fig. \ref{fig:classificationcase}, d). This is an important global context, the summarization of the attribution of all instances.

The current projection basis is depicted as the width of a bar, the feature contributions to the horizontal positions. The radial tour is going to vary one selected feature to a full contribution, zero contribution, and back to the attribution projection. Doing so tests a feature's sensitivity to the structure identified by the local explanation. The default feature selected has the largest discrepancy between the attribution of the PI and CI. The data (scaled by standard deviation away from the mean) is projected through the current basis --- the horizonal position of Fig. \ref{fig:classificationcase} e and Fig. \ref{fig:regressioncase} d.

The following sections elaborate on the difference in displays from applying this approach in classification and regression tasks. Now that we have introduced the global view and corresponding radial tour, let us discuss the differences between the classification and regression cases.

@@

\caption{Overview of the cheem viewer for classification tasks. Global view inputs, (a), set the PI, CI, and color statistic. Global view, (b) PC1 by PC2 for data space, attribution space, and prediction by observed y (visual of the confusion matrix for classification). Points are colored by predicted class, and red circles indicate misclassified instances. Radial tour inputs (c) select features to include, which feature is changed in the tour, and the presence of the parallel coordiate lines connecting attribution distribution. The visual depicting 1D basis (d) shows the distribution of the feature attribution, bars show the current projection with the black bar is being varied. Panel (e) is the resulting projection of the data shown as density in the classification case.}\label{fig:classificationcase}
\end{figure}

@@

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to substitute for the overly-used iris data and is quite similar in form. After removing incomplete instances, there are 333 instances and we will use the four physical measurements, \texttt{bill\_length\_mm} (\texttt{b\_l}), \texttt{bill\_depth\_mm} (\texttt{b\_d}), \texttt{flipper\_length\_mm} (\texttt{f\_l}), \texttt{body\_mass\_g} (\texttt{wgt}), for this illustration. A random forest model was fit with species as the response feature.

@@

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identified by SHAP in the (wrong) prediction for this instance is mostly \texttt{b\_l} and \texttt{b\_d} with small contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap, that is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this, we use the radial tour to vary the contribution of flipper length (\texttt{f\_l}). (In our exploration, this was the third feature explored. It is typically useful to explore the features with larger contributions, here \texttt{b\_l} and \texttt{b\_d}, but when doing this, nothing was revealed about how the PI differed from other Gentoos). On varying \texttt{f\_l} as it contributes more to the projection (b, right), we see that more, and more, this penguin looks like a Chinstrap. This suggests that \texttt{f\_l} should be considered an important feature for explaining the (wrong) prediction.

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{f\_l}) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two features, mostly because it has an unusually low value of flipper length relative to other Gentoos.

@@

The chocolates dataset consists of 88 instances of ten nutritional measurements determined from their labels and labeled as either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University class STAT503 from nutritional information from the manufacturer's website, and normalized to 100g equivalents. The data is available in the \texttt{cheem} package. A random forest model is used for the classification of chocolate type.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate that is nutritionally more like milk should not be considered a healthy alternative. It is interesting to explore which of the nutritional features contribute most to misclassification.

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove four features with the lowest contribution for the selected instance and vary the contribution of \texttt{fiber}. The misclassification of two instances seems improbable when sugar is near the max contribution. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolates}
\end{figure}

This type of exploration is shown in Figure \ref{fig:casechocolates}, where a chocolate labeled dark but predicted to be milk is chosen as the PI (instance 22). It is compared with a CI that is a correctly classified dark chocolate (instance 7). The PCA plot, and the SHAP PCA plots (a) show a big difference between the two chocolate types but with confusion for a handful of instances. The misclassifications are clearer in the observed vs predicted plot, and can be seen to be mistaken in both ways: milk to dark and dark to milk.

The attribution projection for chocolate 22 suggests that Fiber, Sugars and Calories are most responsible for its incorrect prediction. The way to read this plot is to see that Fiber has a large negative value, while Sugars and Calories have reasonably large positive values. In the density plot, instances on the very left of the display would have high values of Fiber (matching the negative projection coefficient) and low values of Sugars and Calories. The opposite would be the interpretation of a point with high values in this plot. The dark chocolates (orange) are mostly on the left, and this is a reason why they are considered to be healthier: high fiber and low sugar. The density for milk chocolates is further to the right, indicating that they generally have low fiber and high sugar.

The instance of interest (dashed line) can be viewed against the comparison instance (dotted line). Now one needs to pay different attention to the parallel plot of the SHAP values, which are local to a particular instance, and the density plot, which is the same projection of all instances as specified by the SHAP values of the instance of interest.

We can quickly compare the feature contributions to the two different predictions from the parallel coordinate plot. The instance of interest differs with the comparison primarily on the Fiber feature, which suggests that this is the reason for the incorrect prediction.

From the density plot, which is the attribution projection corresponding to the instance of interest, both instances are more like dark chocolates. If we vary the contribution of Sugars, and completely remove Sugars from the projection, this is where the difference becomes apparent. When primarily Fiber is examined, instance 22 looks more like a milk chocolate.
